---
title: "GSEAlens: Preparing Input Data from limma-voom and DESeq2 Workflows"
author: "Dudali"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{GSEAlens: Preparing Input Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```



# Installation



GSEAlens is available on Bioconductor. Install the stable release with:



```{r install-package, eval=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))

    install.packages("BiocManager")

BiocManager::install("GSEAlens")

```



The development version can be installed from GitHub:



```{r install-github, eval=FALSE}

if (!requireNamespace("pak", quietly = TRUE))

    install.packages("pak")

pak::pkg_install("DDL095/GSEAlens")

```



# Introduction



This vignette describes how to prepare the input objects that `r Biocpkg("GSEAlens")`

consumes. **GSEAlens itself does not perform differential expression analysis (DEG)**;

it accepts already-fitted `MArrayLM` objects from `r Biocpkg("limma")` or `DESeqDataSet`

objects from `r Biocpkg("DESeq2")`. The following steps are based on the standard

workflows of those packages and are provided here for completeness.



For full DEG workflow documentation, please consult:



- `r Biocpkg("limma")` and `r Biocpkg("edgeR")` vignettes (limma-voom pipeline)

- `r Biocpkg("DESeq2")` vignette

- `r Biocpkg("airway")` package for the example dataset



# Example Data



We use the `r Biocpkg("airway")` dataset (4 vs 4 samples, dex treatment).



```{r setup-environment, results='hide'}

library(airway)

data(airway)

expression_data <- airway

```



# limma-voom Workflow



GSEAlens requires a **no-intercept design** (`~0+group`) so that column names

directly correspond to group names, enabling precise contrast construction.



```{r limma-voom-data-preparation}

library(edgeR)

library(limma)

group_level <- expression_data$dex

design <- model.matrix(~0+group_level)

colnames(design) <- levels(group_level)

compare_end <- combn(levels(group_level), 2, simplify = FALSE)

contrast_strings <- sapply(compare_end, function(x) paste(x[2], x[1], sep = " - "))

contrast_matrix <- makeContrasts(contrasts = contrast_strings, levels = design)



genes_df <- data.frame(

  gene_id = SummarizedExperiment::rowData(expression_data)$gene_id,

  symbol = SummarizedExperiment::rowData(expression_data)$symbol,

  gene_biotype = SummarizedExperiment::rowData(expression_data)$gene_biotype

)

genes_df$Length <- SummarizedExperiment::rowData(expression_data)$gene_seq_end -

                   SummarizedExperiment::rowData(expression_data)$gene_seq_start + 1



gsea_limma_voom_data <- edgeR::DGEList(

  counts = SummarizedExperiment::assay(expression_data, "counts"),

  genes = genes_df,

  norm.factors = NULL,

  group = group_level,

  remove.zeros = TRUE

)



# Filter low-expression genes, keep protein-coding RNAs, deduplicate symbols

total_counts <- edgeR::cpm(gsea_limma_voom_data) |> rowSums()

dup_symbols <- gsea_limma_voom_data$genes$symbol[duplicated(gsea_limma_voom_data$genes$symbol)]

keep <- rep(TRUE, nrow(gsea_limma_voom_data))

for (gene in dup_symbols) {

  idx <- which(gsea_limma_voom_data$genes$symbol == gene)

  best_idx <- idx[which.max(total_counts[idx])]

  remove_idx <- idx[idx != best_idx]

  keep[remove_idx] <- FALSE

}

gsea_limma_voom_data <- gsea_limma_voom_data[keep, ]

rownames(gsea_limma_voom_data) <- gsea_limma_voom_data$genes$symbol



keep_biotype <- gsea_limma_voom_data$genes$gene_biotype == "protein_coding"

gsea_limma_voom_data <- gsea_limma_voom_data[keep_biotype, ]

gsea_limma_voom_data <- edgeR::normLibSizes(gsea_limma_voom_data, method = "TMM")

isexpr <- rowSums(edgeR::cpm(gsea_limma_voom_data) > 1) >= 3

gsea_limma_voom_data <- gsea_limma_voom_data[isexpr, ]

```



Perform limma-voom fitting to obtain the `fit` object.



```{r limma-voom-processing}

VoomOutPut <- voom(gsea_limma_voom_data, design)

fit <- lmFit(object = VoomOutPut, design = design) |>

  contrasts.fit(contrasts = contrast_matrix) |>

  eBayes()

```



# DESeq2 Workflow (SummarizedExperiment Input)



```{r keep-protein-coding-rna}

library("DESeq2")

dds_se <- DESeqDataSet(expression_data, design = ~ cell + dex)

gene_biotypes <- SummarizedExperiment::rowData(dds_se)$gene_biotype

keep_protein_coding <- gene_biotypes == "protein_coding"

dds_se <- dds_se[keep_protein_coding, ]

```



```{r remove-low-expression-rna}

smallestGroupSize <- 3

keep <- rowSums(DESeq2::counts(dds_se) >= 10) >= smallestGroupSize

dds_se <- dds_se[keep, ]

rownames(dds_se) <- SummarizedExperiment::rowData(dds_se)$gene_name



total_counts <- rowSums(SummarizedExperiment::assay(dds_se))

dup_genes <- rownames(dds_se)[duplicated(rownames(dds_se))]

keep <- rep(TRUE, nrow(dds_se))

for (gene in dup_genes) {

  idx <- which(rownames(dds_se) == gene)

  best_idx <- idx[which.max(total_counts[idx])]

  remove_idx <- idx[idx != best_idx]

  keep[remove_idx] <- FALSE

}

dds_se <- dds_se[keep, ]

```



```{r run-deseq2, results='hide'}

dds_se <- DESeq(dds_se)

```



# DESeq2 Workflow (Count Matrix Input)



```{r matrix-modification}

DDS_rawdata <- expression_data

gene_biotypes <- SummarizedExperiment::rowData(DDS_rawdata)$gene_biotype

keep_protein_coding <- gene_biotypes == "protein_coding"

DDS_rawdata <- DDS_rawdata[keep_protein_coding, ]



smallestGroupSize <- 3

keep_epd <- rowSums(SummarizedExperiment::assay(DDS_rawdata, "counts") >= 10) >= smallestGroupSize

DDS_rawdata <- DDS_rawdata[keep_epd, ]

rownames(DDS_rawdata) <- SummarizedExperiment::rowData(DDS_rawdata)$gene_name



total_counts <- rowSums(SummarizedExperiment::assay(DDS_rawdata))

keep_name <- rep(TRUE, nrow(DDS_rawdata))

dup_genes <- rownames(DDS_rawdata)[duplicated(rownames(DDS_rawdata))]

for (gene in dup_genes) {

  idx <- which(rownames(DDS_rawdata) == gene)

  best_idx <- idx[which.max(total_counts[idx])]

  remove_idx <- idx[idx != best_idx]

  keep_name[remove_idx] <- FALSE

}

DDS_rawdata <- DDS_rawdata[keep_name, ]

```



```{r matrix-extraction}

cts <- SummarizedExperiment::assay(DDS_rawdata, "counts")

coldata <- as.data.frame(SummarizedExperiment::colData(DDS_rawdata))

coldata <- coldata[, c("cell", "dex")]

coldata$cell <- factor(coldata$cell)

coldata$dex <- factor(coldata$dex)

```



```{r create-dds-object, results='hide'}

dds <- DESeqDataSetFromMatrix(countData = cts, colData = coldata, design = ~ dex)

dds <- DESeq(dds)

```



# Next Step



Once `fit`, `dds_se`, and `dds` are prepared, return to the main vignette:



```r

vignette("GSEAlens")

```



```{r sessionInfo}

sessionInfo()

```

