---
title: "GSEAlens: An Interactive Exploration Platform for Gene Set Enrichment Analysis"
author: "Dudali"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{GSEAlens: An Interactive Exploration Platform for Gene Set Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to the R Package

**GSEAlens** derives its name from **Lens**, symbolizing a magnifying glass that helps researchers delve deeper into key pathways revealed by GSEA (Gene Set Enrichment Analysis).

GSEAlens provides a web-based interactive platform for displaying pathway introductions and descriptions, integrating AI-assisted pathway enrichment result export functionality. By encapsulating workflows and standardizing input formats, this R package simplifies the viewing and exploration of GSEA enrichment analysis results.

## Installation Instructions

```{r}
# Install using devtools
if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
}
devtools::install_github("DDL095/GSEAlens")
```

# Example Workflow

This example uses the `airway` package, which contains two groups with four replicates each.

```{r 环境准备}
library(GSEAlens)
# First check if airway is installed
if (!requireNamespace("airway", quietly = TRUE)) {
  # If not installed, check and install BiocManager first
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  # Install airway via BiocManager
  BiocManager::install("airway")
}
library("airway")
data(airway)
expression_data <- airway
```

## Input Data Preparation

### limma-voom Workflow

Following the standard workflow, first use the edgeR package to construct the DGEList object, design a no-intercept matrix and rename row names, then create the contrast matrix for fitting, filter low-expression genes, retain protein-coding RNA, and use symbol as row names.

***Note: GSEAlens is based on a no-intercept design limma-voom workflow (~0+group) to ensure that column names directly correspond to group names.***

```{r limma-voom 数据准备}
library(edgeR)
library(limma)
group_level <- airway@colData@listData[["dex"]] 
design <- model.matrix(~0+group_level)
colnames(design) <- levels(group_level)
compare_end <- combn(levels(group_level), 2, simplify = FALSE)
contrast_strings <- sapply(compare_end, function(x) paste(x[2], x[1], sep = " - ")) 
contrast_matrix <- makeContrasts(contrasts = contrast_strings, levels = design)

genes_df <- data.frame(
  gene_id = airway@rowRanges@elementMetadata@listData$gene_id,
  symbol = airway@rowRanges@elementMetadata@listData$symbol,
  gene_biotype = airway@rowRanges@elementMetadata@listData$gene_biotype
)
genes_df$Length <- airway@rowRanges@elementMetadata@listData$gene_seq_end - 
                   airway@rowRanges@elementMetadata@listData$gene_seq_start + 1
gsea_limma_voom_data <- edgeR::DGEList(
  counts = assay(airway, "counts"),
  genes = genes_df,
  norm.factors = NULL,
  group = group_level,
  remove.zeros = T
)

total_counts <- edgeR::cpm(gsea_limma_voom_data) |> rowSums()  

dup_symbols <- gsea_limma_voom_data$genes$symbol[duplicated(gsea_limma_voom_data$genes$symbol)]

keep <- rep(TRUE, nrow(gsea_limma_voom_data))

for (gene in dup_symbols) {
  idx <- which(gsea_limma_voom_data$genes$symbol == gene)  
  best_idx <- idx[which.max(total_counts[idx])]            
  remove_idx <- idx[idx != best_idx]                       
  keep[remove_idx] <- FALSE
}

gsea_limma_voom_data <- gsea_limma_voom_data[keep, ]

rownames(gsea_limma_voom_data) <- gsea_limma_voom_data$genes$symbol

keep_biotype <- gsea_limma_voom_data$genes$gene_biotype == "protein_coding"
gsea_limma_voom_data <- gsea_limma_voom_data[keep_biotype,]

gsea_limma_voom_data <- edgeR::normLibSizes(gsea_limma_voom_data, method = "TMM")

isexpr <- rowSums(edgeR::cpm(gsea_limma_voom_data) > 1) >= 3
gsea_limma_voom_data <- gsea_limma_voom_data[isexpr,]
```

Perform limma-voom fitting to obtain the `fit` object.

```{r limma-voom 处理}
VoomOutPut <- voom(gsea_limma_voom_data, design)

fit <- lmFit(object = VoomOutPut, design = design) %>% 
  contrasts.fit(fit = ., contrasts = contrast_matrix) %>% 
  eBayes(.)
```

### DESeq2 Workflow

#### SummarizedExperiment Input

Refer to the DESeq2 package documentation for data cleaning and processing. First, remove non-coding RNA.

```{r 仅保留编码蛋白质的RNA}
library("DESeq2")
dds_se <- DESeqDataSet(expression_data, design = ~ cell + dex)
gene_biotypes <- rowData(dds_se)$gene_biotype
keep_protein_coding <- gene_biotypes == "protein_coding"
dds_se <- dds_se[keep_protein_coding, ]
dds_se
```

Then remove low-expression genes and use symbol as row names.

```{r 去除低表达的RNA}
smallestGroupSize <- 3
keep <- rowSums(counts(dds_se) >= 10) >= smallestGroupSize
dds_se <- dds_se[keep,]
dds_se
rownames(dds_se) <- dds_se@rowRanges@elementMetadata@listData[["gene_name"]]

total_counts <- rowSums(SummarizedExperiment::assay(dds_se))
keep <- rep(TRUE, nrow(dds_se))
dup_genes <- rownames(dds_se)[duplicated(rownames(dds_se))]
for (gene in dup_genes) {
  idx <- which(rownames(dds_se) == gene)  
  best_idx <- idx[which.max(total_counts[idx])]  
  remove_idx <- idx[idx != best_idx]  
  keep[remove_idx] <- FALSE
}
dds_se <- dds_se[keep, ]
```

Run DESeq2 processing.

```{r 运行DESeq2}
dds_se <- DESeq(dds_se)
```

### Count Matrix Input

Filter low-expression genes, retain only protein-coding RNA, and rename matrix row names to symbol.

```{r 矩阵修订}
DDS_rawdata <- expression_data

gene_biotypes_epd <- rowData(DDS_rawdata)$gene_biotype
keep_protein_coding_epd <- gene_biotypes == "protein_coding"
DDS_rawdata <- DDS_rawdata[keep_protein_coding, ]

smallestGroupSize <- 3
keep_epd <- rowSums(assay(DDS_rawdata, "counts") >= 10) >= smallestGroupSize
DDS_rawdata <- DDS_rawdata[keep_epd,]

rownames(DDS_rawdata) <- DDS_rawdata@rowRanges@elementMetadata@listData[["gene_name"]]

total_counts <- rowSums(SummarizedExperiment::assay(DDS_rawdata))
keep_name <- rep(TRUE, nrow(DDS_rawdata))
dup_genes <- rownames(DDS_rawdata)[duplicated(rownames(DDS_rawdata))]
for (gene in dup_genes) {
  idx <- which(rownames(DDS_rawdata) == gene)  # All row indices for this gene
  best_idx <- idx[which.max(total_counts[idx])]  # Row with highest expression
  remove_idx <- idx[idx != best_idx]  # Rows to remove (same gene name but not highest expression)
  keep_name[remove_idx] <- FALSE
}
DDS_rawdata <- DDS_rawdata[keep_name, ]
```

Obtain matrix information.

```{r 矩阵提取}
cts <- assay(DDS_rawdata,"counts")
coldata <- DDS_rawdata@colData %>% as.data.frame.array()
coldata <- coldata[,c("cell","dex")]
coldata$cell <- factor(coldata$cell)
coldata$dex <- factor(coldata$dex)
```

Assemble into DDS object and process.

```{r}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ dex)
dds <- DESeq(dds)
```

## GSEAlens Processing

### Creating GSEA Pathway Object

Use the `build_gsea_pathways` function to construct the pathway object for GSEA enrichment analysis.

```{r}
gsea_pathwaysets <- GSEAlens::build_gsea_pathways(species = "HS",auto_select = c(17,18,19,20,26))
```

### Assembling the Computation Object

Through the `setup_gsea_env` function, assemble the `GSEAEnv` object for computational analysis. Different workflows use the same function with different input data. For the limma-voom workflow, since the `fit` object does not contain the original gene read counts, the filtered `DGEList` object used to generate the `fit` object must be additionally included, which in this example is the multi-step filtered `gsea_limma_voom_data`.

Include limma-voom workflow object for analysis.

```{r}
gseadata_limmavoom <- GSEAlens::setup_gsea_env(fit = fit,pathway_obj = gsea_pathwaysets,expr_data = gsea_limma_voom_data)
```

Include DESeq2 SummarizedExperiment object for analysis.

```{r}
gseadata_se <- GSEAlens::setup_gsea_env(fit = dds_se,pathway_obj = gsea_pathwaysets)
```

Include DESeq2 Count matrix workflow object for analysis.

```{r}
gseadata_dds <- GSEAlens::setup_gsea_env(fit = dds,pathway_obj = gsea_pathwaysets)
```

### Running Processing

All objects are processed using the `batch_calc_gsea` function with no difference.

Parallel computing note: Adjust the workers option based on your computer's performance to set the number of cores for computation. More comparisons require higher core settings for better efficiency.

```{r}
# limma-voom workflow
gsea_res_limmavoom <- GSEAlens::batch_calc_gsea(gseadata_limmavoom, 
                                                 custom_series_name = "limmavoom_data", 
                                                 workers = 4,  # Adjust based on number of comparisons and computer performance
                                                 force = TRUE)

# DESeq2 SummarizedExperiment workflow
gsea_res_se <- GSEAlens::batch_calc_gsea(gseadata_se, 
                                          custom_series_name = "dds_se_data", 
                                          workers = 4,
                                          force = TRUE)

# DESeq2 Count matrix workflow
gsea_res_dds <- GSEAlens::batch_calc_gsea(gseadata_dds, 
                                           custom_series_name = "dds_data", 
                                           workers = 4,
                                           force = TRUE)
```

### Interactive Analysis and Viewing

After running `batch_calc_gsea`, an RDS file will be generated in the corresponding directory. Load this file, or use `import_gsea_capsule` to read it. The `import_gsea_capsule` function automatically organizes related files into the folder of your RMD file or R script when reading, and provides a data inspection overview.

```{r}
gsea_res <- import_gsea_capsule("/path/to/your/files/")
# Or read the RDS file directly
# gsea_res <- readRDS("/path/of/your/file/")
```

Once the data is loaded, you can use `launch_gsea_app` for interactive viewing.

```{r}
launch_gsea_app(gsea_res)
```

# Package Intermediate Object Documentation

## GseaEnv Object

The `GseaEnv` object returned by the `setup_gsea_env` function contains the following components:

| Component | Description |
|-----------|-------------|
| `backend_info` | Backend type information (limma-voom or DESeq2) |
| `contrast_registry` | Contrast registry containing all pairwise comparison information |
| `de_store` | Differential expression analysis result storage |
| `expr_bundle` | Expression data bundle (raw counts, normalized matrix, sample metadata) |
| `geneset` | Geneset information (TERM2GENE, metadata dictionary, species) |

## GseaRes Object

The `GseaRes` object returned by the `batch_calc_gsea` function contains the following components:

| Component | Description |
|-----------|-------------|
| `metadata` | Computation metadata (runtime, cores used, parameter settings) |
| `backend_info` | Backend type information |
| `contrast_registry` | Contrast registry |
| `de_store` | Differential expression analysis result storage |
| `expr_bundle` | Expression data bundle |
| `geneset_info` | Geneset information |
| `results` | GSEA result list, one entry per contrast |

## GseaTask Object

The `GseaTask` object returned by the `extract_gsea_task` function is used for single-contrast analysis:

| Component | Description |
|-----------|-------------|
| `gsea_res` | GSEA result object |
| `meta` | Metadata (contrast information, geneset name, expression data) |

# Session Information

```{r sessionInfo}
sessionInfo()
```

# References
